Program Equivalence : A Relational Separation Logic interactive prover implemented in Maude

### 2018-06-30 paper/paper_after_guide.tex
proof-read the paper after modifications

\tableofcontents
\pagebreak
\begin{abstract}
	Lucrarea de față descrie procesul de dezvoltare al unui utilitar interactiv utilizat în analiza formală a programelor, folosindu-se de Logica Separatoare Relațională \cite{relational} și Logica Separatoare \cite{primer} \cite{SeparationLogic}, utilizând Maude \cite{maudesite} drept cadru de dezvoltare. Componenta interactivă se îmbină cu o componentă automatizată, prin care părți din demonstrație vor fi realizate automat. Lucrarea este orientata pe rezolvarea problemei prin intermediul funcționalităților oferite de Maude și va include secvențe de cod ce vor ilustra conceptele cheie legate de modelarea si aplicarea fundamentelor teoretice utilizate pe parcursul proiectului.
\end{abstract}
\section*{Introduction}
In the present paper I describe the development of an interactive tool for reasoning about how two programs are related, in a formal manner. \\

The tool is an implementation of Hoare Logic, which allows formal reasoning about a program along with two of its extensions, namely the Separation Logic and Relational Separation Logic \cite{relational}. The two extensions simplify the Hoare Logic proofs, mainly using the "*" connector, allowing for local reasoning about the effects of statements in a program. I have the tool implemented in Maude, a high performance logical framework with powerful metalanguage applications that facilitate the implementation of executable environments for logics.\\

@@

The paper is organized as follows: 
\begin{itemize}
	\item {\textbf{Section 1} describes the main personal contributions to the realization of this project;}
	\item {\textbf{Section 2} presents the problem this project aims to solve;}
	\item {\textbf{Section 3} shortly presents some other projects related to this one;}
	\item {\textbf{Section 4} briefly touches upon the theoretical foundations of this project, together with the technologies used to implement them;}
	\item {\textbf{Section 5}, the main section of this paper, presents my approach to solving the problem, and discuss how it can be improved in the future.}
\end{itemize}
\section{Contributions}
Personal contributions:
\begin{itemize}
	\item Modelling the Relational Separation Logic and Separation Logic using Maude equational and rewriting logic specifications; 
	\item Developing an interactive tool for reasoning about program behaviour using the aforementioned logics;
	\item Automation of tasks that make the tool more convenient to use;
	\item Example of formal proof performed using the tool - Appendix \ref{proofExample}
\end{itemize}
Throughout the development of the prover, I have learned how to use a different, unfamiliar technology, that is a good fit for  modelling theoretical concepts which were also new to me. This learning by applying process properly introduced me to both the technology and the logics involved in formally verifying program equivalence. 
\section{Description of the Problem}
Comparing programs or code fragments and studying their equivalence is part of the activities of every software engineer, when they are testing an alternative implementation for an existing solution, fixing bugs, launching new product versions, etc. Naturally, for every process completed manually there are efforts being made in order to make it more efficient, less error-prone and, in the end, automate the process all together. These goals can be achieved by creating an easy to use and reliable formal  method of verifying program equivalence.
\\

@@

The problem this project aims to solve is reducing the gap between theoretical and practical aspects of formal proofs for program equivalence by creating an executable environment for Relational Separation Logic in which the user can prove the equivalence of programs. Through the creation of such a tool, the theoretical concepts become more accessible, easier to apply and understand. Making formal verification methods more accessible leads to an overall increase in the quality of software, by replacing the existing methods of verifying equivalence, that are often prone to errors, with more robust ones.\\

The difficult aspects of creating this tool include:
\begin{itemize}
	\item Losing as little generality as possible when modelling the theoretical concepts in a programming language;
	\item Approaching computationally-hard problems required for Relational Separation Logic proofs;
	\item Understanding the complexities of Relational Separation Logic and create a user experience which facilitate the application of the logic.
\end{itemize}

@@

Good references for reading on separation logic include: 
\begin{itemize}
	\item \cite{primer} - for a good introduction and some interesting usages ;
	\item \cite{SeparationLogic} - for a more in depth explanation, including the mathematical fundaments of the logic.
\end{itemize} 
\subsection{Relational Separation Logic}
Relational Separation Logic, which is the central concept of our project builds upon Separation Logic to reason about how two programs are related, by using the concept of a \textbf{Hoare Quadruple}:
\[\{R\}\ {C_1 \atop C_2}\   \{T\},\]
where \(R\) represents a \textbf{relation} between the two program states holding before the execution of commands \(C_1\) and \(C_2\), while \(T\) represents a \textbf{relation} that holds after the execution of the two commands. 
\\

@@

{
	The semantics of the previously are standard, but a few clarifications are necessary to point out how our language constructs interact with the storage model, as well as a few differences from regular programming languages:

@@

		\item{Verbosity of operators - most operators are replaced by their literal names (\(+\) becomes \(\mathtt{plus}\) etc.). I have opted for this approach to avoid syntactic complication when implementing the language in Maude, because most of the operators are already defined in Maude for its built-in types, and overloading them could cause conflicts since we are basing our own defined types on Maude's primitives.}

@@

		\subitem{The rest of the operators are self-explanatory}

@@

	\paragraph{Maude Modelling of the Language}

@@

Each proof rule in the Separation Logic and Relational Separation Logic has been translated into a Maude rewrite rule. The prover is centered around the concept of \textbf{Goals} which can be one of the three things: 
\begin{itemize}
	\item {Hoare Quadruple} - corresponding to Relational Separation Logic;
	\item {Hoare Triple} - corresponding to Separation Logic;
	\item {Implication between relations and assertions}.
\end{itemize} 
The proof rules are interpreted in a bottom-up manner, meaning that, applying a rule to a goal, which must match the conclusion of the used rule, will generate new goals consisting of the hypothesis: Listing \ref{slConseq} presents how a Separation Logic rule has been translated into Maude syntax and Listing \ref{rslConseq} presents the same process for the Relational Separation Logic counterpart of the rule.
\begin{flushleft}
	The Separation Logic rule of \textsf{Consequence}
\end{flushleft}
	\medskip
	\[
	\begin{prooftree}
	\Hypo{P \Rightarrow P\prime}
	\Hypo{\{P\prime\} C \{Q\prime\}}
	\Hypo{Q\prime \Rightarrow Q}
	\Infer 3 {\{P\}\  C  \{Q\}}
	\end{prooftree}	
	\]
	Is represented in Maude syntax as: 
\begin{lstlisting}[caption=Separation Logic Consequence rule,label=slConseq]
	rl [SeparationLogic-Consequence] : {P} C {Q} => ((P => P1) <> ({P1} C {Q1})) <> (Q1 => Q) [nonexec] .
\end{lstlisting}
\begin{flushleft}
	The Relational Separation Logic rule of \textsf{Consequence}
\end{flushleft}
	\[
	\begin{prooftree}
	\Hypo{R \Rightarrow R_1}
	\Hypo{\hq{R_1}{C}{C\prime}{S_1}}
	\Hypo{S_1 \Rightarrow S}
	\Infer 3 {\hq{R}{C}{C\prime}{S}}
	\end{prooftree}	
	\]
	is represented in Maude syntax as: 
	\begin{lstlisting}[caption=Relational Separation Logic Consequence rule,label=rslConseq]
	rl [Consequence] : { R } C1 -- C2  { S } => ((R => R1) <> ({R1} C1 -- C2 {S1})) <> (S1 => S) [nonexec] .
	\end{lstlisting}
The \texttt{<>} operator is used to separate sub-goals generated by the rule.
\\

The \texttt{---} operator is used to separate the two snippets of code mentioned in the \textbf{Hoare Quadruple}.
\\

All the rules are given the attribute \texttt{[nonexec]} because we want them to be applied in a controlled manner, at the meta-level. Some of the rules, including the examples, are not even valid executable Maude rules, because they make use of variables before binding them For example, in the Separation Logic rule of \textsf{Consequence}, \texttt{P1} is used in a rule before it is bound. At the meta-level, however we can manually bind values to those variables, as we will show when discussing how we apply these rules to goals.
\subsubsection{Sort Hierarchy}
Much in the same manner as we modelled the language grammar, Relational Separation Logic and Separation Logic have been modelled by defining sorts and declaring operators using these sorts. The main point of reference for our modelling has been \cite{relational}, for both logics. Some examples of translations from specification to code are included below.
\begin{itemize}
	\item \texttt{Assertion} is the sort used to handle Separation Logic assertions. It is a super sort of the \texttt{BExp} sort defined as part of our language;
	\item \texttt{HoareTriple} is the sort for Separation Logic \textbf{Hoare Triples} (Listing \ref{HT}):
	\begin{lstlisting}[caption=Hoare Triple constructor operator,label=HT]
op {_}_{_} : Assertion Command Assertion -> HoareTriple .\end{lstlisting}
	\item \texttt{Relation} is the sort for Relational Separation Logic relations. As with \texttt{Assertion}, it is a super sort of the \texttt{BExp} sort of our language. Because both the \texttt{Relation} and \texttt{Assertion} are super sorts of the same sort, Maude will display a few warnings which are not affecting functionality, but in the future, creating a separate \texttt{BExp} sort for both the \texttt{Assertion} and \texttt{Relation} sorts would be a better option.
	\item \texttt{HoareQuadruple} denotes the concept with the same name, in the context of Relational Separation Logic .\\
	Listing \ref{HQ} language constructs relevant for Relational Separation Logic.
	\begin{figure}[h]
	\begin{lstlisting}[caption=Examples of Relational Separation Logic specific constructs,label=HQ]
--- Operator which creates a relation from two assertions
op [_;_] : Assertion Assertion -> Relation .
--- Operator which constructs a Hoare Quadruple 
op {_}_--_{_} : Relation Command Command Relation -> HoareQuadruple .\end{lstlisting}
	\end{figure}
\end{itemize}
The other operators for the Assertion and Relation languages have been defined in the same manner as the programming language operators, following their presentation in \cite{relational}.

@@

We will make a distinction between \textbf{Loop state}, which is specific to Maude and the \textbf{Prover state}, which is specific to our use-case, and holds the necessary data related to the state of the proof we are currently working on. The \textbf{Prover state} is encapsulated within \textbf{Loop state} (see Figure \ref{fig:states}).
\begin{figure}[h]
	\includegraphics[width=\textwidth]{"pictures/States".png}
	\caption{Loop and Prover State structure}
	\label{fig:states}
\end{figure}
\subsection{State of the Prover}
Following the examples in the Maude Manual \cite{manual}, our prover state (Figure \ref{fig:states}) is composed of multiple sections modified accordingly during the execution of the prover:
\begin{itemize}
	\item{\texttt{Action} - represents the current action to be executed by the prover;}
	\item{\texttt{Root Goal} - represents the end-goal of the proof;}
	\item{\texttt{Current Goal} - the goal being treated currently - all actions will be applied to this goal;}
	\item{\texttt{Goal Stack} - stack of goals necessary to be proven in order to prove the root goal. The current goal is the last element popped from the stack and new elements are generated and pushed by the application of proof rules to the current goal;}
	\item{\texttt{Proven Goals} - list of goals that have already been proven, or axioms. Each goal will be first checked to see if it matches any goal in this list to prevent repeating the same proofs;}
	\item{\texttt{Staging Output} - contains the output generated by the current action, before being passed to the Loop state. This staging mechanism enables each rewriting rule to generate output independent of the others.}
\end{itemize}
\subsection{Input and Output Handling}
To understand how our prover handles input and output, it is important to note that the flow of the LOOP-MODE module is as follows: 
\begin{enumerate}
	\item User input is placed in the \(\mathtt{INPUT}\) member of the loop state;
	\item Maude engine applies all the possible rewritings, resulting in a new loop state. All the logic of the prover is implemented here, in the form of rewriting rules which modify the prover state;
	\item The member \(\mathtt{OUTPUT}\) of the new loop state is taken and displayed at the console.
\end{enumerate}
There are two main rules used for handling input and output: 
\begin{enumerate}
	\item{The \texttt{in} rule (Listing \ref{inRule}), which is responsible for parsing user input and modifying the action in the prover state accordingly;}
	\begin{lstlisting}[caption=The \texttt{in} rule,label=inRule]
crl [in] : 
--- [ INPUT, <PROVER_STATE>, OUTPUT]
[QIL, < noRule ; RG ; G ; GS ; GL ; nil > , QIL']
=> 
if T:ResultPair? :: ResultPair  
--- If parsing succeeded, place the parsed action in
--- the prover state
then [nil, < downTerm(getTerm(T:ResultPair?), noAction) ; RG ; G ; GS ; GL ; nil >, QIL']  
--- If the parsing failed, display an error
else [nil, < noRule ; RG ; G ; GS ; GL ; nil >, 'ERROR QIL]  
fi  
--- Try to parse the input as an term of sort Action
if QIL =/= nil /\ T:ResultPair? := metaParse(upModule('PROVER-GRAMMAR, false), QIL, 'Action) .
\end{lstlisting}
	\item{The \texttt{out} rule (Listing \ref{outRule}), which takes the staging output from the prover state and appends it to the currently existing output in the loop state.}
	\begin{lstlisting}[caption=The \texttt{out} rule,label=outRule]
crl [out] :
[QIL, < A ; RG ; G ; GS ; GL ;  QIL' >, QIL'']
=> [QIL, < A ; RG ; G ; GS ; GL ; nil >, QIL''  QIL']
if QIL' =/= nil .
\end{lstlisting}
\end{enumerate}

\subsection{Prover Execution Flow}
This section describes the flow of a Relational Separation Logic proof using our prover. The most important parts of the execution of the prover are controlled through the first member of a prover state, \texttt{Action}, which specifies the next step in the processing of the current goal. There are two types of actions, \texttt{System Actions}, colored in the figure in purple and \texttt{User Actions}, colored in light-yellow. The diagram showing how the prover works is presented in Figure \ref{fig:ProverDiagram}.
\\
\begin{enumerate}
	\item The user issues the command \texttt{loop init .} in the Maude console. This will initialize the loop state;
	\item Now the user will have to specify the goal it wants to prove, by issuing the command \begin{center}(\texttt{prove(G)}),\end{center} where G is the representation of the goal using the syntax defined for our logics. The command will populate the required fields of the prover state (\texttt{RootGoal}, \texttt{GoalStack}) with G.
	\\

@@

			\item The Consequence rule is applied at a meta level, using the \texttt{metaXapply} operator [line 11] from the \texttt{META-LEVEL} module. To this operator we also pass the values of the parameters (in their meta representation) with which to bind variables appearing in the \texttt{Consequence} rule;
			\item The result is then checked for errors to see if there were some errors when applying the \texttt{metaXapply} operator (line 8) or if the term returned is not actually of the required type (line 3);

@@

			\item The last thing to note is an example of how the rules generate output independent of other applications. In the last field of the \texttt{Prover State}, as \texttt{StagingOutput}, there will be a list of quoted identifiers which will eventually get moved to the loop state and then outputted to the user.

@@

\subsubsection{Variable substitution}

@@

	However, Maude does not allow for variables to appear for the first time on the right hand side of a rewrite rule which needs to be used without meta-level computations (problem present when modelling the Separation Logic as well). We had to find a workaround for this problem and the solution we came up with was to write the axiom in its meta-representation, and then use the \texttt{downTerm} operator to bring it down in the meta-hierarchy (Listing \ref{axiom}).

	\begin{lstlisting}[label=axiom,caption=Initialization of Prover state]
rl [proveGoal] : < prove(HQ) ; RG ; G ; GS ; GL ; nil > =>
< noAction ; HQ ; noGoal ; HQ <> emptyGL ; 
downTerm('`{_`}_`{_`}['_|->_['E:AExp,'_:AExp],'`[_`]:=_['E:AExp,'F:AExp],'_|->_['E:AExp,'F:AExp]] , emptyGL)
--- ... [other axioms omitted for clarity] 
; 'begin '\n > .\end{lstlisting}
Before attempting to match terms we also interpret the equality between variables using the \texttt{replace} operator (Listing \ref{replaceSymb}) introduced previously to reduce the \texttt{Goal} to a simpler form, which might match to an axiom (Listing \ref{rewriteBeforeMatch}).
	\begin{lstlisting}[caption=Equation reducing the goal before attempting to match it,label=rewriteBeforeMatch]
ceq anyMatches(G, G1 <> GL) = G1
---if G := G1 .
if metaXmatch(upModule('PROVER-INTERFACE, false), 
upTerm(G1),
--- Try to reduce G as much as possible before matching
getTerm(metaRewrite(upModule('SYMBOLIC-EXECUTION, false), upTerm(G), unbounded)),
nil, 0, unbounded, 0) :: MatchPair .
	\end{lstlisting} 
	\begin{lstlisting}[caption=Rule intepreting variable equality,label=replaceSymb]
rl {P * (emp /\ X eqs Y)} X1 := E {Q} =>  
if X1 =/= X then 
{replace(P, X, Y)} replace(X1 := E, X, Y) {replace(Q, X, Y)}
else 
{replace(P, X, Y)} X1 := replace(E, X, Y) {Q}
fi .\end{lstlisting}

\subsubsection{Automatic Proof of Implications}
\label{autoImpl}
Automatic proof of implications is one of the most challenging aspects of this paper and this section will present our incipient approach to this problem.

@@

The second parameter controls the depth up to which the search goes. The prover will try automatically to search a solution within a depth of 3 rewrites, but the user can manually specify searching for a solution using the \texttt{auto x} command, where \texttt{x} is the depth up to which Maude should try to search the solution (our experiments show that searches up to a depth around 12 should finish use a reasonable amount of resources and finish in under 10 minutes - this observation of course depends on the implication and the number of possible rewrites starting from it). The found path is presented to the user if the implication has been proven. 
\\

Almost all the implications in Appendix \ref{proofExample} can be automatically proven, the exceptions being the ones for which the running time of the prover is too high, since they require a depth higher than 15 - those are manually admitted using the \texttt{ok} command.
\\

>>>>

\tableofcontents
\pagebreak
\begin{abstract}
	Lucrarea de față descrie procesul de dezvoltare al unui utilitar interactiv utilizat în analiza formală a echivalenței programelor, folosindu-se de Logica Separatoare Relațională \cite{relational} și Logica Separatoare \cite{primer} \cite{SeparationLogic}, utilizând Maude \cite{maudesite} drept cadru de dezvoltare. Componenta interactivă se îmbină cu o componentă automatizată, prin care, părți din demonstrație vor fi realizate automat. Lucrarea este orientată pe rezolvarea problemei prin intermediul funcționalităților oferite de Maude și va include secvențe de cod ce vor ilustra conceptele cheie legate de modelarea și aplicarea fundamentelor teoretice utilizate pe parcursul proiectului.
\end{abstract}
\section*{Introduction}
In the present paper I describe the development of an interactive tool for reasoning about how two programs are related, in a formal manner. \\

The tool is an implementation of Hoare Logic, which allows formal reasoning about a program, along with two of its extensions, namely the Separation Logic and Relational Separation Logic \cite{relational}. The two extensions simplify the Hoare Logic proofs, mainly using the "*" connector, allowing for local reasoning about the effects of statements in a program. I have implemented the tool in Maude, a high performance logical framework with powerful metalanguage applications that facilitate the implementation of executable environments for logics.\\

@@

The paper is organized as follows: 
\begin{itemize}
	\item {\textbf{Section 1} describes the main personal contributions to the realization of this project;}
	\item {\textbf{Section 2} presents the problem this project aims to solve;}
	\item {\textbf{Section 3} shortly presents some other projects related to this one;}
	\item {\textbf{Section 4} briefly touches upon the theoretical foundations of this project, together with the technologies used to implement them;}
	\item {\textbf{Section 5}, the main section of this paper, presents my approach to solving the problem, and discuss how it can be improved in the future.}
\end{itemize}
\section{Contributions}
Personal contributions to:
\begin{itemize}
	\item Modelling the Relational Separation Logic and Separation Logic using Maude equational and rewriting logic specifications; 
	\item Developing an interactive tool for reasoning about program behaviour using the aforementioned logics;
	\item Automation of tasks that make the tool more convenient to use;
	\item Example of formal proof performed using the tool - Appendix \ref{proofExample}
\end{itemize}
Throughout the development of the prover, I have learned how to use a different, unfamiliar technology, that is a good fit for  modelling theoretical concepts, which were also new to me. This learning by applying process properly introduced me to both the technology and the logics involved in formally verifying program equivalence. 
\section{Description of the Problem}
Comparing programs or code fragments and studying their equivalence is part of the activities of every software engineer, when they are testing an alternative implementation for an existing solution, fixing bugs, launching new product versions, etc. Naturally, for every process completed manually there are efforts being made in order to make it more efficient, less error-prone and, in the end, automate the process all together. These goals can be achieved by creating an easy to use and reliable formal  method of verifying program equivalence.
\\

@@

The problem this project aims to solve is reducing the gap between theoretical and practical aspects of formal proofs for program equivalence, by creating an executable environment for Relational Separation Logic in which the user can prove the equivalence of programs. Through the creation of such a tool, the theoretical concepts become more accessible, easier to apply and understand. Making formal verification methods more accessible leads to an overall increase in the quality of software, by replacing the existing methods of verifying equivalence, that are often prone to errors, with more robust ones.\\

The difficult aspects of creating this tool include:
\begin{itemize}
	\item Modelling the theoretical concepts in a programming language;
	\item Approaching computationally-hard problems required for Relational Separation Logic proofs;
	\item Understanding the complexities of Relational Separation Logic and creating an user experience that facilitates the application of the logic.
\end{itemize}

@@

Good references for reading on separation logic include: 
\begin{itemize}
	\item \cite{primer} - for a good introduction and some interesting usages;
	\item \cite{SeparationLogic} - for a more in depth explanation, including the mathematical fundaments of the logic.
\end{itemize} 
\subsection{Relational Separation Logic}
Relational Separation Logic, which is the central concept of our project builds upon Separation Logic to reason about how two programs are related, by using the concept of a \textbf{Hoare Quadruple}:
\[\{R\}\ {C_1 \atop C_2}\   \{T\},\]
where \(R\) represents a \textbf{relation} between the two program states holding before the execution of commands \(C_1\) and \(C_2\), while \(T\) represents a \textbf{relation} that holds after the execution of the two commands. 
\\

@@

	The semantics of the operators are standard, but a few clarifications are necessary to point out how our language constructs interact with the storage model, as well as a few differences from regular programming languages:

@@

		\item{Verbosity of operators - most operators are replaced by their literal names (\(+\) becomes \(\mathtt{plus}\) etc.). I have opted for this approach to avoid syntactic complication when implementing the language in Maude, because most of the operators are already defined in Maude for its built-in types, and overloading them could cause conflicts since we are basing our own defined types on the primitives of Maude.}

@@

		\subitem{The rest of the operators are self-explanatory.}

@@

	\paragraph{Maude Modelling of the Language:}

@@

Each proof rule in the Separation Logic and Relational Separation Logic has been translated into a Maude rewrite rule. The prover is centered around the concept of \textbf{Goals} which can be one of the three things: 
\begin{itemize}
	\item {Hoare Quadruple} - corresponding to Relational Separation Logic;
	\item {Hoare Triple} - corresponding to Separation Logic;
	\item {Implication between relations and assertions}.
\end{itemize} 
The proof rules are interpreted in a bottom-up manner, meaning that, applying a rule to a goal, which must match the conclusion of the used rule, will generate new goals consisting of the hypothesis: Listing \ref{slConseq} presents how a Separation Logic rule has been translated into Maude syntax and Listing \ref{rslConseq} presents the same process for the Relational Separation Logic counterpart of the rule.
\begin{flushleft}
	The Separation Logic rule of \textsf{Consequence}:
\end{flushleft}
	\medskip
	\[
	\begin{prooftree}
	\Hypo{P \Rightarrow P\prime}
	\Hypo{\{P\prime\} C \{Q\prime\}}
	\Hypo{Q\prime \Rightarrow Q}
	\Infer 3 {\{P\}\  C  \{Q\}}
	\end{prooftree}	
	\]
	Is represented in Maude syntax as: 
\begin{lstlisting}[caption=Separation Logic Consequence rule,label=slConseq]
	rl [SeparationLogic-Consequence] : {P} C {Q} => ((P => P1) <> ({P1} C {Q1})) <> (Q1 => Q) [nonexec] .
\end{lstlisting}
\begin{flushleft}
	The Relational Separation Logic rule of \textsf{Consequence}:
\end{flushleft}
	\[
	\begin{prooftree}
	\Hypo{R \Rightarrow R_1}
	\Hypo{\hq{R_1}{C}{C\prime}{S_1}}
	\Hypo{S_1 \Rightarrow S}
	\Infer 3 {\hq{R}{C}{C\prime}{S}}
	\end{prooftree}	
	\]
	is represented in Maude syntax as: 
	\begin{lstlisting}[caption=Relational Separation Logic Consequence rule,label=rslConseq]
	rl [Consequence] : { R } C1 -- C2  { S } => ((R => R1) <> ({R1} C1 -- C2 {S1})) <> (S1 => S) [nonexec] .
	\end{lstlisting}
The \texttt{<>} operator is used to separate sub-goals generated by the rule.
\\
The \texttt{---} operator is used to separate the two snippets of code mentioned in the \textbf{Hoare Quadruple}.
\\

All the rules are given the attribute \texttt{[nonexec]} because we want them to be applied in a controlled manner, at the meta-level. Some of the rules, including the examples, are not even valid executable Maude rules, because they make use of variables before binding them For example, in the Separation Logic rule of \textsf{Consequence}, \texttt{P1} is used in a rule before it is bound. At the meta-level, however we can manually bind values to those variables, as we will show when discussing how we apply these rules to goals.
\subsubsection{Sort Hierarchy}
Much in the same manner as we modelled the language grammar, Relational Separation Logic and Separation Logic have been modelled by defining sorts and declaring operators using these sorts. The main point of reference for our modelling has been \cite{relational}, for both logics. Some examples of translations from specification to code are included below.
\begin{itemize}
	\item \texttt{Assertion} is the sort used to handle Separation Logic assertions. It is a super sort of the \texttt{BExp} sort defined as part of our language;
	\item \texttt{HoareTriple} is the sort for Separation Logic \textbf{Hoare Triples} (Listing \ref{HT}):
	\begin{lstlisting}[caption=Hoare Triple constructor operator,label=HT]
op {_}_{_} : Assertion Command Assertion -> HoareTriple .\end{lstlisting}
	\item \texttt{Relation} is the sort for Relational Separation Logic relations. As with \texttt{Assertion}, it is a super sort of the \texttt{BExp} sort of our language. Because both the \texttt{Relation} and \texttt{Assertion} are super sorts of the same sort, Maude will display a few warnings which are not affecting functionality, but in the future, creating a separate \texttt{BExp} sort for both the \texttt{Assertion} and \texttt{Relation} sorts would be a better option;
	\item \texttt{HoareQuadruple} denotes the concept with the same name, in the context of Relational Separation Logic .\\
	Listing \ref{HQ} presents language constructs relevant for Relational Separation Logic.
	\begin{figure}[h]
	\begin{lstlisting}[caption=Examples of Relational Separation Logic specific constructs,label=HQ]
--- Operator which creates a relation from two assertions
op [_;_] : Assertion Assertion -> Relation .
--- Operator which constructs a Hoare Quadruple 
op {_}_--_{_} : Relation Command Command Relation -> HoareQuadruple .\end{lstlisting}
	\end{figure}
\end{itemize}
The other operators for the Assertion and Relation languages have been defined in the same manner as the programming language operators, following their presentation in \cite{relational}.

@@

We will make a distinction between \textbf{Loop state}, which is specific to Maude, and the \textbf{Prover state}, which is specific to our use-case, and holds the necessary data related to the state of the proof we are currently working on. The \textbf{Prover state} is encapsulated within \textbf{Loop state} (see Figure \ref{fig:states}).
\begin{figure}[h]
	\includegraphics[width=\textwidth]{"pictures/States".png}
	\caption{Loop and Prover State structure}
	\label{fig:states}
\end{figure}
\subsection{State of the Prover}
Following the examples in the Maude Manual \cite{manual}, our prover state (Figure \ref{fig:states}) is composed of multiple sections modified accordingly during the execution of the prover:
\begin{itemize}
	\item{\texttt{Action} - represents the current action to be executed by the prover;}
	\item{\texttt{Root Goal} - represents the end-goal of the proof;}
	\item{\texttt{Current Goal} - the goal being treated currently - all actions will be applied to this goal;}
	\item{\texttt{Goal Stack} - stack of goals necessary to be proven in order to prove the root goal. The current goal is the last element popped from the stack and new elements are generated and pushed by the application of proof rules to the current goal;}
	\item{\texttt{Proven Goals} - list of goals that have already been proven, or axioms. Each goal will first be checked to see if it matches any goal in this list to prevent repeating the same proofs;}
	\item{\texttt{Staging Output} - contains the output generated by the current action, before being passed to the Loop state. This staging mechanism enables each rewriting rule to generate output independent of the others.}
\end{itemize}
\subsection{Input and Output Handling}
To understand how our prover handles input and output, it is important to note that the flow of the LOOP-MODE module is as follows: 
\begin{enumerate}
	\item User input is placed in the \(\mathtt{INPUT}\) member of the loop state;
	\item Maude applies all the possible rewritings, resulting in a new loop state. All the logic of the prover is implemented here, in the form of rewriting rules which modify the prover state;
	\item The \(\mathtt{OUTPUT}\) member of the loop state is taken and displayed at the console.
\end{enumerate}
There are two main rules used for handling input and output: 
\begin{enumerate}
	\item{The \texttt{in} rule (Listing \ref{inRule}), which is responsible for parsing user input and modifying the action in the prover state accordingly;}
	\begin{lstlisting}[caption=The \texttt{in} rule,label=inRule]
crl [in] : 
--- [ INPUT, <PROVER_STATE>, OUTPUT]
[QIL, < noRule ; RG ; G ; GS ; GL ; nil > , QIL']
=> 
if T:ResultPair? :: ResultPair  
--- If parsing succeeded, place the parsed action in
--- the prover state
then [nil, < downTerm(getTerm(T:ResultPair?), noAction) ; RG ; G ; GS ; GL ; nil >, QIL']  
--- If the parsing failed, display an error
else [nil, < noRule ; RG ; G ; GS ; GL ; nil >, 'ERROR QIL]  
fi  
--- Try to parse the input as an term of sort Action
if QIL =/= nil /\ T:ResultPair? := metaParse(upModule('PROVER-GRAMMAR, false), QIL, 'Action) .
\end{lstlisting}
	\item{The \texttt{out} rule (Listing \ref{outRule}), which takes the staging output from the prover state and appends it to the currently existing output in the loop state.}
	\begin{lstlisting}[caption=The \texttt{out} rule,label=outRule]
crl [out] :
[QIL, < A ; RG ; G ; GS ; GL ;  QIL' >, QIL'']
=> [QIL, < A ; RG ; G ; GS ; GL ; nil >, QIL''  QIL']
if QIL' =/= nil .
\end{lstlisting}
\end{enumerate}

\subsection{Prover Execution Flow}
This section describes the flow of a Relational Separation Logic proof using our prover. The most important parts of the execution of the prover are controlled through the first member of a prover state, \texttt{Action}, which specifies the next step in the processing of the current goal. There are two types of actions: \texttt{System Actions} and \texttt{User Actions}. The diagram showing how the prover works is presented in Figure \ref{fig:ProverDiagram}.
\\
\begin{enumerate}
	\item The user issues the command \texttt{loop init .} in the Maude console. This will initialize the loop state;
	\item Now the user will have to specify the goal it wants to prove, by issuing the command \begin{center}(\texttt{prove(G)}),\end{center} where G is the representation of the goal using the syntax defined for our logics. The command will populate the required fields of the prover state (\texttt{RootGoal}, \texttt{GoalStack}) with G.
	\\

@@

			\item The Consequence rule is applied at the meta level, using the \texttt{metaXapply} operator (line 11) from the \texttt{META-LEVEL} module. To this operator we also pass the values of the parameters (in their meta representation) with which to bind variables appearing in the \texttt{Consequence} rule;
			\item The result is then checked for errors to see if the \texttt{metaXapply} operator has not been applied correctly (line 8), or if the term returned is not actually of the required type (line 3);

@@

			\item The last thing to note is an example of how the rules generate output independently of others. In the last field of the \texttt{Prover State}, as \texttt{StagingOutput}, there will be a list of quoted identifiers which will eventually get moved to the loop state and then outputted to the user.

@@

\subsubsection{Variable Substitution}

@@

	However, Maude does not allow for variables to appear for the first time on the right hand side of a rewrite rule which needs to be used without meta-level computations (problem present when modelling Separation Logic as well). We had to find a workaround for this problem and the solution we came up with was to write the axiom in its meta-representation, and then use the \texttt{downTerm} operator to bring it down in the meta-hierarchy (Listing \ref{axiom}).

	\begin{lstlisting}[label=axiom,caption=Initialization of Prover state]
rl [proveGoal] : < prove(HQ) ; RG ; G ; GS ; GL ; nil > =>
< noAction ; HQ ; noGoal ; HQ <> emptyGL ; 
downTerm('`{_`}_`{_`}['_|->_['E:AExp,'_:AExp],'`[_`]:=_['E:AExp,'F:AExp],'_|->_['E:AExp,'F:AExp]] , emptyGL)
--- ... [other axioms omitted for clarity] 
; 'begin '\n > .\end{lstlisting}
Before attempting to match terms we also interpret the equality between variables using the \texttt{replace} operator (Listing \ref{replaceSymb}) introduced previously, to reduce the \texttt{Goal} to a simpler form, which might match to an axiom (Listing \ref{rewriteBeforeMatch}).
	\begin{lstlisting}[caption=Equation reducing the goal before attempting to match it,label=rewriteBeforeMatch]
ceq anyMatches(G, G1 <> GL) = G1
---if G := G1 .
if metaXmatch(upModule('PROVER-INTERFACE, false), 
upTerm(G1),
--- Try to reduce G as much as possible before matching
getTerm(metaRewrite(upModule('SYMBOLIC-EXECUTION, false), upTerm(G), unbounded)),
nil, 0, unbounded, 0) :: MatchPair .
	\end{lstlisting} 
	\begin{lstlisting}[caption=Rule intepreting variable equality,label=replaceSymb]
rl {P * (emp /\ X eqs Y)} X1 := E {Q} =>  
if X1 =/= X then 
{replace(P, X, Y)} replace(X1 := E, X, Y) {replace(Q, X, Y)}
else 
{replace(P, X, Y)} X1 := replace(E, X, Y) {Q}
fi .\end{lstlisting}

\subsubsection{Automatic Proof of Implications}
\label{autoImpl}
Automatic proof of implications is one of the most challenging aspects of this project and this section will present our incipient approach to this problem.

@@

The second parameter in Listing \ref{eqImplicationProving} - line 1, controls the depth up to which the search goes. The prover will try automatically to search a solution within a depth of 3 rewrites, but the user can manually specify searching for a solution using the \texttt{auto x} command, where \texttt{x} is the depth up to which Maude should try to search the solution (our experiments show that searches up to a depth around 12 should finish use a reasonable amount of resources and finish in under 10 minutes - this observation of course depends on the implication and the number of possible rewrites starting from it). The found path is presented to the user if the implication has been proven. 
\\

Almost all the implications in Appendix \ref{proofExample} can be automatically proven, the exceptions being the ones for which the compute resources required by the prover are too high, since they require a depth higher than 15 - those are manually admitted using the \texttt{ok} command.
\\
